EvoSwarm: A Cloud-Native Architecture for the Reproducible Execution of Multi-population Optimization Algorithms

### 2020-08-21 draft.tex
Update draft.tex

\usepackage{lineno,hyperref}
\usepackage{rotating}
\usepackage{algpseudocode,algorithm}
\usepackage{hyperref}
\usepackage{listings}
\usepackage{xcolor}
\usepackage[normalem]{ulem}
\useunder{\uline}{\ul}{}

@@

%
\begin{figure}[h!tbp]
  \centering
  \includegraphics[width=\textwidth]{secs_per_instance}
  \caption{Boxplot of the median time required for the execution of 30 
  instances of function $f_4$, in seconds, different number of workers and populations sizes. The y-axis is shown in log scale.  }
  \label{fig:spworker:time}
\end{figure}
%
\begin{table}[tbp]
  \caption{The median of the time required to finish an instance of the function $f_4$, 30 runs
  with different number of workers and populations}
  \label{tab:time}
  \vspace{0.25cm}
    \centering
    \begin{tabular}{ccrrr}
    \hline
              &         & 5 Populations    & 10 Populations & 20 Populations \\
     Dimension& Workers & Median of rate & Median of rate & Median of rate  \\
    \hline
          10  & 1w      & 22.24  & 20.48           & 24.55  \\
              & 2w      & 10.31  & 10.00           & 12.79 \\
              & 4w      &  6.23  & 4.75            &  5.70 \\
              & 8w      &  5.86  &  2.33           & 2.74 \\
              & 16w     &  5.96  &  2.33           & \textbf{1.59}\\

    \hline
          20  & 1w   & 18.36   & 65.18  & 80.31 \\
              & 2w   & 34.18   & 33.46  & 38.87 \\
              & 4w   & 19.55   & 16.58  & 18.69 \\
              & 8w   & 18.38   & 7.77   & 8.66 \\
              & 16w  & 18.36   & 7.27   & \textbf{4.57} \\
    \hline
  \end{tabular}
\end{table}

@@

% I think this is now a bit redundant or we need to move it to the parent section - M
% But is a good intro to the results.
% OK - JJ

%In principle, the main issue is how to find the right number
%of workers for a particular problem, but in general, we wanted to check the
%communication overhead when the size of the problem increases and how the
%presented framework reacted to it.
% Commenting because we did not checked the communication over head - M

Figure \ref{fig:spworker:time} shows the median (30 runs) of the execution time achieved by 2, 4 , 8, and 16  
workers, to finish an instance of the function $f_4$, for dimensions 10 and 20, with 5, 10, and 20
population sizes.

Each box-plot indicates the time required for the execution of 30 function
instances, in seconds. Each row has a different log scale on the y-axis, to
compensate for the dimension increments. It is important to notice that the
execution time can also be affected by the algorithm finding a solution with an
error smaller than the threshold, before completing all the FEs.

@@

However, when we compare the two options, we find some subtle differences: When
using one or two workers, using 5 populations takes less time in most cases. In
Table~\ref{tab:time} results with p-value of the two-tailed Wilcoxon
signed-ranks test was grater than the level of significance of 0.05 (i.e.,
accepting the null hypothesis of inequality) are marked in boldface. A reason
for this could be that when having more populations, the migration of
individuals takes more time from the beginning to reach to all of them. For
instance, when a single worker has 10 populations, only after processing every
one of them, it will start receiving mixed populations. Although, in all cases,
and even with a single worker, populations are heterogeneous, individuals
resulting from the mixing of populations take more time to propagate. With four
workers, almost the same results can be achieved regardless of our selection
between five or ten populations. Also, there is not much difference between 4
and 8 workers, when using five populations, which might indicate that some
workers are idling while populations are being processed by other workers. In
general, we can observe that keeping a ratio between the number of workers and
populations a bit over 1:1 exploits better the computing resources available.
This also decreases the relation between the overhead needed to migrate and
process incoming and outgoing population and the total computing time: more
populations will need more time to do that.

@@

\begin{figure}[h!tbp]
  \centering
  \includegraphics[width=\textwidth]{evals_per_instance}
  \caption{Boxplot of the number of function evaluations required to reach the target, for 30 instances of function $f_4$, with different number of workers and populations sizes}
  \label{fig:spworker:evals}
\end{figure}

>>>>

\usepackage{lineno,hyperref}
\usepackage{rotating}
\usepackage{algpseudocode,algorithm}
\usepackage{hyperref}
\usepackage{listings}
\usepackage{xcolor}
\usepackage{multirow}
\usepackage[normalem]{ulem}
\useunder{\uline}{\ul}{}

@@

\begin{figure}[h!tbp]
  \centering
  \includegraphics[width=\textwidth]{evals_per_instance}
  \caption{Boxplot of the number of function evaluations required to reach the target, for 30 instances of function $f_4$, with different number of workers and populations sizes}
  \label{fig:evals_instance}
\end{figure}

\begin{table}[tbp]
  \caption{The median of the number of evaluations required to finish an instance of the function $f_4$, 30 runs
  with different number of workers and populations}
  \label{tab:evals_instance}
  \vspace{0.25cm}
  \centering

  \begin{tabular}{|c|c|r|r|r|}
  \hline
  \multicolumn{1}{|c}{} &   \multicolumn{1}{c}{}  & \multicolumn{3}{|c|}{Number of Populations} \\
  \hline
  Dimension           & Workers & \multicolumn{1}{c|}{5}       & \multicolumn{1}{c|}{10}      & \multicolumn{1}{c|}{20}          \\
  \hline
  \multirow{5}{*}{10} & 1w      & 159,625      & \textbf{142,368*}      & 151,680*     \\
                      & 2w      & 160,258      & \textbf{137,385}      & 150,445*     \\
                      & 4w      & 161,144      & \textbf{129,031}      & 144,354     \\
                      & 8w      & 159,893      & \textbf{{\ul 127,987}}      & 141,896     \\
                      & 16w     & 167,678      & \textbf{137,446}      & 138,852     \\
                      \hline
  \multirow{5}{*}{20} & 1w      & 462,069      & \textbf{396,841}      & 421,249*     \\
                      & 2w      & 437,952      & \textbf{395,814}      & 407,984     \\
                      & 4w      & 437,563      & \textbf{396,587}      & 408,461     \\
                      & 8w      & 440,130      & \textbf{370,264}      & 372,265     \\
                      & 16w     & 443,108      &         386,272       & \textbf{{\ul 355,999}}\\
                      \hline  
  \end{tabular}
  \end{table}

%
\begin{figure}[h!tbp]
  \centering
  \includegraphics[width=\textwidth]{secs_per_instance}
  \caption{Boxplot of the median time required for the execution of 30 
  instances of function $f_4$, in seconds, different number of workers and populations sizes. The y-axis is shown in log scale.  }
  \label{fig:spworker:time}
\end{figure}
%
\begin{table}[tbp]
  \caption{The median of the time required to finish an instance of the function $f_4$, 30 runs
  with different number of workers and populations}
  \label{tab:time}
  \vspace{0.25cm}
    \centering
    \begin{tabular}{ccrrr}
    \hline
              &         & 5 Populations    & 10 Populations & 20 Populations \\
     Dimension& Workers & Median of Time   & Median of Time & Median of Time  \\
    \hline
          10  & 1w      & 22.24  & 20.48           & 24.55  \\
              & 2w      & 10.31  & \textbf{10.00} (2.05)    & 12.79 \\
              & 4w      &  6.23  & \textbf{4.75} (2.10)     &  5.70 \\
              & 8w      &  5.86  &  \textbf{2.33} (2.03)           & 2.74 \\
              & 16w     &  5.96*  &  2.33*           & {\ul 1.59} \\

    \hline
          20  & 1w   & 73.87   & 65.18                 & 80.31 \\
              & 2w   & 34.18   & 33.46                 & 38.87 \\
              & 4w   & 19.55   & \textbf{16.58} (2.02) & 18.69 \\
              & 8w   & 18.38   & \textbf{7.77} (2.15)  & 8.66 \\
              & 16w  & 18.36*  & 7.27                  & {\ul 4.57} \\
    \hline
  \end{tabular}
\end{table}

@@

In the previous section, we have shown that as we increase the number of
populations and workers, we increase the ratio of the number of evaluations per
second in the same proportion. However, it is vital to notice that we can also
affect the number of evaluations we need to execute to find a solution. This
reduction means that the multi-population based algorithm can perform better
when we increase the number of populations, or in some cases, even workers, as the experiments
show. If we increase the evaluations per second and, at the same time, decrease
the number of evaluations needed, we can have a superlinear speedup of the
algorithm, if we measure the time required to complete an instance.

First, we will see how the configuration we selected can reduce the number of
evaluations needed in an experiment. Figure \ref{fig:evals_instance} shows the
median (30 runs) of the number of evaluations needed to complete an instance of
the function $f_4$. For every selection in the number of populations, we could
expect to have about the same number of FEs needed regardless of the number of
workers. However, we can see that for some configurations, i.e., 10D and ten
populations, eight workers will need less FEs then one (p = 0.012). Also, for
20D and twenty populations, there are differences between one and four workers
(p = 0.08) and between four and sixteen (p = 0.01). These results indicate that
there are benefits in the potential change of order in the message queue
resulting from the parallelization of work.

In Table \ref{tab:evals_instance} we can see some additional insights in more
detail. When we increase the number of populations from five to ten, or twenty,
the algorithm needs fewer FEs in both cases. The reduction is more pronounced
when using ten populations, needing fewer FEs in all but a single case (shown in
boldface). In some cases, marked with an asterisk, there is no statistical
difference between the number of evaluations; this happens when using only one
or two workers. The configurations that required less FEs have an underline in
the table.

The reduction in the number of FEs has an impact when we measure the speed up
concerning the time needed to complete an experiment. In this case, we have
measured the median of the seconds needed to complete an instance of the
problem. Figure \ref{fig:spworker:time} shows in each box-plot the median (30
runs) of the execution time achieved by 2, 4, 8, and 16 workers, to finish an
instance of the function $f_4$, for dimensions 10 and 20, when using 5, 10, and
20 population sizes. We present each dimension in a row having a log scale on
the y-axis, to compensate for the dimension increments and scalability.

@@

However, when we compare the two options, we find some subtle differences: When
using one or two workers, using 5 populations takes less time in most cases. In
Table~\ref{tab:time} results with p-value of the two-tailed Wilcoxon
signed-ranks test was grater than the level of significance of 0.05 (i.e.,
accepting the null hypothesis of inequality) are marked in boldface. A reason
for this could be that when having more populations, the migration of
individuals takes more time from the bdyeginning to reach to all of them. For
instance, when a single worker has 10 populations, only after processing every
one of them, it will start receiving mixed populations. Although, in all cases,
and even with a single worker, populations are heterogeneous, individuals
resulting from the mixing of populations take more time to propagate. With four
workers, almost the same results can be achieved regardless of our selection
between five or ten populations. Also, there is not much difference between 4
and 8 workers, when using five populations, which might indicate that some
workers are idling while populations are being processed by other workers. In
general, we can observe that keeping a ratio between the number of workers and
populations a bit over 1:1 exploits better the computing resources available.
This also decreases the relation between the overhead needed to migrate and
process incoming and outgoing population and the total computing time: more
populations will need more time to do that.

@@
